---
title: "Some useful miscellaneous functions that are not in the other vignettes"
author: "Andrew Edwards"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Some useful miscellaneous functions that are not in the other vignettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "Last rendered on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r, instr, echo = FALSE, eval = FALSE}
# To build either run this line or click knit button in RStudio
rmarkdown::render("misc-sizeSpectraFit.Rmd")
```

```{r, load, include = FALSE}
load_all()
# library(sizeSpectraFit)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " ",
  fig.width = 7,
  fig.height = 6
)
```

This vignette for the [sizeSpectraFit R
package](https://github.com/andrew-edwards/sizeSpectraFit?tab=readme-ov-file#sizespectrafit)
contains some extra miscellaneous functions from the package, that are not
covered in the other vignettes (or doing so would add extra clutter to those
vignettes) but that might be useful for users.

There are still further functions in the package (called by other functions)
that users are not likely to need, but all code for them is in the `R/`
directory for the package at
https://github.com/andrew-edwards/sizeSpectraFit/tree/main/R. To list all exported functions you
can type `getNamespaceExports("sizeSpectraFit")`; each will have a help
function.

## Note about customised functions 

Note that we have customised plotting functions that automatically get applied
to, for example, the results of `size_spectrum_fit()` depending on the class of the
argument. In case you are unaware of how this works in R, here is a simple
annotated example.
```{r, class, fig.height = 5.5}
res <- fit_size_spectrum(sim_vec)      # Fit a simulated vector of values
class(res)                             # Check the class
plot(res)
```
Now, because `res` has a specific class, namely `"size_spectrum_numeric" "list"`, R will look at
the first element of the class when we call `plot()`, and see if there is a
specific plotting function for that element, in the form of
`plot.size_spectrum_numeric()` in this case. This then gets used in the above
plot. Similar behaviour occurs for other types of results and some other functions. This is
useful to know for digging into the help files of such functions (you need to
know how to figure out the explicit function name).

Okay, so here are the extra the functions that are not mentioned in the other
vignettes. See the help files for full
details, here I just highlight them briefly to bring them to your attention
(and to remind myself that they exist!).

## PLB distribution functions: `dPLB`, `dPLB_agg`, etc.

For the bounded power-law (PLB) distribution, standard R functions are available
to give the probability density and distribution functions, and random number
generation.

These are in the standard R format like `dnorm()` for the normal
distribution. So we have:

  * `dPLB()`: probability density function

  * `pPLB()`: cumulative distribution function $P(X \leq x)$ 

  * `rPLB()`: random generation of values

  * `qPLB()`: quantile function
  
as described in [Edwards et
al. (2017)](http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12641/full). Random
generation uses the inverse method (e.g. p1215 of Edwards 2008, Journal of
Animal Ecology, 77:1212-1222). `qPLB()` is from inverting the cumulative
distribution function. Unbounded power-law distributions are also included (as
`dPL()` etc.) for completeness (except for `qPL()`) but are likely not needed.

For aggregated bounded power-law distributions across multiple samples, we
include `dPLB_agg()` and `pPLB_agg()`, based on mathematics worked out
in Appendix B of [Quevedo et al. (2026)](https://www.sciencedirect.com/science/article/pii/S2351989426001769). Results then feed into `plot_aggregate()` to generate the plot.
See the [fit-aggregated
vignette](https://andrew-edwards.github.io/sizeSpectraFit/vignettes/fit-aggregated.html)
for details, and the help files.

## Calculations: `calc_mle_conf()` 

This is called from the fitting functions to calculate the maximum likelihood
estimate of a parameter and its 95% confidence interval by using the profile
log-likelihood method described in the Ecological Detective (Hilborn and
Mangel, 1997, page 162). It can be used for any log-likelihood functions. Our
fitting functions specify the correct log-likelihood function to be used;
e.g. `fit_size_spectrum.numeric()`  uses `neg_ll_mle_method()` which gives the
exact formula for the negative log-likelihood of a power-law distribution given
a set of numerical measurements.

Fits are done using `nlm()`, and warnings can be suppressed with `suppress_warnings =
TRUE` (sometimes there are many, but these all seem to be 

```{r, eval = FALSE}
Warning in nlm(f = negll.fn, p = p, ...) : 

NA/Inf replaced by maximum positive value.
```

This is due to the likelihood function blowing up, presumably when searching
some very very unlikely region of parameter space.

For the
confidence interval it calls our `calc_confidence_interval()` function. If the interval reaches one of the bounds, then
`calc_mle_conf()` repeats the calculation with a larger interval. 

## Detecting and removing outliers

`detect_outliers()` uses the appropriate function depending on the class
of the argument, one of: 
`detect_outliers.size_spectrum_mlebins()`,
`detect_outliers.determine_xmin_and_fit_mlebins()`, or
`detect_outliers.remove_outliers_mlebins()`. 
It analyses gaps to look for outliers, namely large values with gaps from
remaining continuous values, from using the MLEbins method.

If already run through `remove_outliers.size_spectrum_mlebins()` then it will
use `detect_outliers.remove_outliers.mlebins()` and can keep iterating
that. See `?remove_outliers` for further details and example code.

If there is a gap between roughly-continuous body sizes and a larger group
then it is useful to test if removing the larger group affects the results.

Used for Sensitivity Example B in Quevedo et al. (2026), see p34 of Supp
Material B. The relevant code is [available
here](https://github.com/andrew-edwards/sizeSpectraFit/tree/main/report/mediterranean/mediterranean-analysis-15).

## Making histogram objects: `make_hist()`, `make_hist_for_binned_counts_mlebin()`, `make_hist_for_binned_counts()`

`make_hist()` converts a vector of body sizes into a histogram list object and
creates 0 counts for missing bins (such that we can automatically use the base R
`plot.histogram()` for plotting). 

`make_hist_for_binned_counts_mlebin()` does similar
for data in a `data.frame` to be used for the MLEbin method (bins are not
overlapping and so can simply be used). 

`make_hist_for_binned_counts()` does
similar for data that has overlapping bins, taking a tibble with `bin_min`, `bin_max`, and `bin_count` and assign counts to new equal bins based on `bin_min`, summing the original counts (which can be non-integer).


## Cumulative biomass functions for plotting

sizeSpectraFit includes functionality to plot normalised biomass style plots
(that people are used to seeing) that are calculated based on our likelihood
calculations (that we have shown to perform better than fitting the regression
of a normalised biomass plot).

`p_biomass()` gives the biomass cumulative distribution function from equations
A.4 and A.8 in [Edwards et
al. (2017)](http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12641/full). This
is the total biomass between `x_min` and the value of `x`, assuming a bounded power-law
distribution of body masses between `x_min` and `x_max` and a given value of
 exponent `b`, and a total of `n` individuals.

This function is used by `p_biomass_bins()` to give total
biomass (and normalised biomass) in each bin. Bins are either created or
prescribed, depending on the input, and uncertainty is calculated if
appropriate. `p_biomass_bins()` works on the results of a call to our main
`fit_size_spectrum()` function, with `p_biomass_bins.size_spectrum_numeric()` or
`p_biomass_bins.size_spectrum_mlebin()` being called as appropriate. See
`?p_biomass_bins()` for explanation.

The output can then be used for plotting LBN biomass type plots. They are used
automatically, along with `plot_lbn_fitted()` in `plot_lbn_style()` which is called if `style = "biomass"`
in the plotting functions. So these functions are getting used behind the scenes
in the call
```{r, eval = FALSE}
plot(res,
     style = "biomass")
```
in the [fit-data.html
vignette](https://andrew-edwards.github.io/sizeSpectraFit/vignettes/fit-data.html).

## Plotting results for multiple groups

`plot_multiple_exponents()` plots the estimated $b$ with confidence intervals for multiple groups and
strata. A somewhat specific plot for the Mediterranean analyses of five species
groups fitted for three strata in Figure B.20 of [Quevedo et al. (2026)](https://www.sciencedirect.com/science/article/pii/S2351989426001769), but
applicable to other sets of results, likely with some tweaking.

The Quevedo results are saved in the package, and the function creates the
figure:
```{r, multexp}
knitr::kable(quevedo_table_b1,
             digits = 2)   # results from Table B.1 of Quevedo et al. (2026)

plot_multiple_exponents(quevedo_table_b1,
                        shade_first = TRUE)
```

## Plotting customisation

`add_ticks()` adds labelled and unlabelled tick marks to a plot (for linear or
log axes) that are calculated automatically if needed, or can be explicitly
specified. See `?add_ticks()`.

## Printing results

Since the raw data are included in the output from
`fit_size_spectrum.numeric()`, the default `print()` would print all data, so we
provide condensed output using a wrapper function
`print.size_spectrum_numeric()` that gets automatically used to show output. It
limits the output to just the first 10 values. To see them all use `print.default()`.

